---
title: "DESeq2 regularization demo plot"
author: "Wolfgang Huber"
output: BiocStyle::html_document
---

This file is for discussing and adapting / updating the code that underlies Figure 8.10 "Shrinkage estimation of logarithmic fold change estimates by use of an empirical prior in DESeq2" in the MSMB book (https://www.huber.embl.de/msmb/Chap-CountData.html#fig:rnaseq-posterior)

# Create the `pasilla` object and call `DESeq` on it

The following code is a summary of what happens in the MSMB book prior to Section 8.10.

```{r loadpkg, message = FALSE}
library("DESeq2")
library("dplyr")
```
```{r createpasilla, cache = TRUE, message = FALSE}
fn = system.file("extdata", "pasilla_gene_counts.tsv",
                  package = "pasilla", mustWork = TRUE)
counts = as.matrix(read.csv(fn, sep = "\t", row.names = "gene_id"))

annotationFile = system.file("extdata",
  "pasilla_sample_annotation.csv",
  package = "pasilla", mustWork = TRUE)
pasillaSampleAnno = readr::read_csv(annotationFile)

pasillaSampleAnno = mutate(pasillaSampleAnno,
  condition = factor(condition, levels = c("untreated", "treated")),
  type = factor(sub("-.*", "", type), levels = c("single", "paired")))

stopifnot(
  !any(is.na(pasillaSampleAnno$condition)),
  !any(is.na(pasillaSampleAnno$type)),
  sum(pasillaSampleAnno$type == "single") == 3,
  sum(pasillaSampleAnno$type == "paired") == 4)

mt = match(colnames(counts), sub("fb$", "", pasillaSampleAnno$file))
stopifnot(!any(is.na(mt)))

pasilla = DESeqDataSetFromMatrix(
  countData = counts,
  colData   = pasillaSampleAnno[mt, ],
  design    = ~ condition)

pasilla = DESeq(pasilla)
```


# Section 8.10.1: Sharing of dispersion information across genes

We already saw an explanation of Bayesian (or empirical Bayes) analysis in
Figure 6.12.  The idea is to use additional information to improve our
estimates (information that we either known a priori, or have from analysis of other, but
similar data). This idea is particularly useful if the data per se are relatively noisy.
`DESeq2` uses an empirical Bayes approach for the estimation of the dispersion
parameters (the $\alpha$s in Equation 8.16) and, optionally, the logarithmic fold changes
(the $\beta$s).  The priors are, in both cases, taken from the distributions of the
maximum-likelihood estimates (MLEs) across all genes. It turns out that both of these
distributions are uni-modal; in the case of the $\beta$s, with a peak at around 0, in the
case of the $\alpha$, at a particular value, the "typical" dispersion. The empirical
Bayes machinery then "shrinks" each per-gene MLE towards that peak, by an amount
that depends on the sharpness of the empirical prior distribution and the precision of the
ML estimate (the better the latter, the less shrinkage will be done). The mathematics are
explained in (Love et al.), and Figure 8.10 visualizes the
approach for the $\beta$s.

```{r shrink1}
library("apeglm")
res1  = results(pasilla, cooksCutoff = FALSE)
res2  = lfcShrink(pasilla, coef = "condition_treated_vs_untreated", res = res1)

# Something like the two lines commented out below may be used to reproduce   
# how Mike Love selected the two genes: they should have similar intercepts,
# large unshrunken fold change and very different Wald statistic (i.e., have
# small / large dispersion, respectively):
#
# with(res1,
#  plot(baseMean, log2FoldChange, log = "x", ylim = c(0, 3), xlim = c(10, 1e5),
#       col = ifelse(padj < 0.1, "red", "black"), cex = log(abs(stat))))
# rownames(res1)[with(res1, identify(baseMean, log2FoldChange))]

genes = c(A = "FBgn0053926", B = "FBgn0260933")
cols  = c(FBgn0053926 = "forestgreen", FBgn0260933 = "dodgerblue3", 
          prior = "black")

df1 = tibble(
  k         = as.vector(counts(pasilla, normalized = TRUE)[genes, ]),
  condition = rep(colData(pasilla)[["condition"]], each = length(genes)),
  gene      = rep(genes, times = ncol(pasilla)))

beta    = seq(from = -1, to = 1.5, length = 500)
kounts  = counts(pasilla)[genes,, drop = FALSE]
disps   = dispersions(pasilla[genes, ]) %>% `names<-`(genes)

sf      = sizeFactors(pasilla)
cond    = as.numeric(pasilla$condition)-1 

priorSigma = sqrt(priorInfo(res2)$prior.control$prior.var)
prior = dnorm(beta, mean = 0, sd = priorSigma)

likelihood = function(k, alpha, intercept) {
  z = vapply(beta, function(b) {
    prod(dnbinom(k, mu = sf * 2^(intercept + b * cond), size = 1/alpha))
  }, numeric(1))
  z / (sum(z) * diff(beta[1:2]))
}

posterior = function(k, alpha, intercept) {
  z = likelihood(k, alpha, intercept) * prior
  z / (sum(z) * diff(beta[1:2]))
}

intercepts = with(mcols(pasilla[genes,]), Intercept) %>% `names<-`(genes)

df2 = bind_rows(
  tibble(beta = beta, y = prior, gene = "prior", what = "pre"),
  bind_rows(
  lapply(genes, function(i) bind_rows(
    tibble(beta = beta, gene = i, what = "pre",
           y = likelihood(k = kounts[i, ], alpha = disps[i],
                          intercept = intercepts[i])),
    tibble(beta = beta, gene = i, what = "post",
           y = posterior(k = kounts[i, ], alpha = disps[i],
                         intercept = intercepts[i]))))
  )
)

is_max = function(y)
  ifelse(seq(along = y) == which.max(y), y, NA_real_)

df2 %<>% group_by(gene, what) %>% mutate(py = is_max(y))

# some consistency checks:
deseqNoPrior = res1[genes, "log2FoldChange"]
deseqPrior   = res2[genes, "log2FoldChange"]
mleFromPlot=c(beta[which.max(likelihood(kounts[1,],disps[1],intercepts[1]))],
              beta[which.max(likelihood(kounts[2,],disps[2],intercepts[2]))])
mapFromPlot=c(beta[which.max( posterior(kounts[1,],disps[1],intercepts[1]))],
              beta[which.max( posterior(kounts[2,],disps[2],intercepts[2]))])
stopifnot(all(abs(deseqNoPrior - mleFromPlot) < .01),
          all(abs(deseqPrior   - mapFromPlot) < .07))
```
```{r shrinkplot1, message = FALSE}
library("ggbeeswarm")
ggplot(df1, aes(x = condition, y = k, col = gene)) + geom_beeswarm(cex = 5) +
      facet_grid(. ~ gene) + ylab("normalized counts") + scale_y_log10() +
      scale_color_manual(values = cols) + theme(legend.position = "none")
```
```{r shrinkplot2, warning = FALSE, fig.cap="Shrinkage estimation of logarithmic fold change estimates by use of an empirical prior in DESeq2. Two genes with similar mean count and MLE logarithmic fold change are highlighted in green and blue. The normalized counts for these genes (upper panel) reveal low dispersion for the gene in blue and high dispersion for the gene in green. In the lower panel, the density plots are shown of the normalized likelihoods (solid lines) and of the posteriors (dashed lines) for the green and blue gene. In addition, the solid black line shows the prior estimated from the MLEs of all genes. Due to the higher dispersion of the green gene, its likelihood is wider and less sharp (indicating less information), and the prior has more influence on its posterior than in the case of the blue gene."}
ggplot(df2, aes(x = beta, col = gene, linetype = what)) +
  geom_line(aes(y = y)) + geom_point(aes(y = py)) +
  scale_color_manual(values = cols) + theme(legend.position = "none") +
  scale_linetype_manual(values = c(pre = "solid", post = "dotted")) +
  xlab(expression(beta)) + ylab("density")
```
